---
title: "Statistical Models"
subtitle: "Lecture 7"
from: markdown+emoji
author: 
    - name: Dr. Silvio Fanzon
      id: sf
      email: S.Fanzon@hull.ac.uk
      url: https://www.silviofanzon.com
      affiliations: University of Hull
    - name: Dr. John Fry
      id: jf
      email: J.M.Fry@hull.ac.uk
      url: https://www.hull.ac.uk/staff-directory/john-fry
      affiliations: University of Hull
---



::: {.content-hidden}
$
{{< include macros.tex >}}
$
:::




# Lecture 7: <br>The chi-squared test{background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Outline of Lecture 7

1. Overview
2. The chi-squared goodness-of-fit test
3. Goodness-of-fit test: Examples
4. Monte Carlo simulations
4. Computational Example 1
5. The chi-squared test of independence/no association
6. Computational Example 2





# Part 1: <br>Overview {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## Problem statement {.smaller}


**Data:** in the form of **numerical counts**

**Test:** difference between observed counts and predictions of theoretical model

**Example**: Blood counts

- Suppose to have counts of blood type for some people

    | A  |  B  |  AB  |  O  |
    |:--:|:---:|:----:|:---:|
    |2162| 738 | 228  |2876 |

- We want to compare the above to the probability model

    | A  |  B  |  AB  |  O  |
    |:--:|:---:|:----:|:---:|
    |1/3 | 1/8 |  1/24| 1/2 |
    




## Counts with multiple factors {.smaller}

**Example:** Relative performance of Man Utd managers

- Each football manager has Win, Draw and Loss count


| Manager | Played  |     Won |  Drawn  | Lost    |
|:-------:|:-------:|:-------:|:-------:|:-------:|
|  Moyes  |    51   |   27    | 9       |   15    |
|Van Gaal | 103     |   54    |  25     |   24    |
| Mourinho|  144    |   84    |  32     |   28    |
|Solskjaer|  168    |    91   |   37    |   40    |


**Questions:**

- Is the number of Wins, Draws and Losses uniformly distributed?
- Are there differences between the performances of each manager?



## Chi-squared test {.smaller}

- Hypothesis testing for counts can be done using chi-squared test

- Chi-squared test is simple to use for real-world project datasets (e.g. dissertations)

- Potentially applicable to a whole range of different models

- Easy to compute by hand/software

- Motivates the more advanced study of contingency tables




# Part 2: <br>Chi-squared <br> goodness-of-fit test {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## Chi-squared goodness-of-fit test {.smaller}

**Categorical Data:** 

- Finite number of possible categories or types
- Observations can only belong to one category
- $O_i$ refers to **observed count** of category $i$

| Type $1$| Type $2$ |  $\ldots$ | Type $n$ |
|:-------:|:--------:|:---------:|:--------:|
|$O_1$    |   $O_2$  | $\ldots$  |    $O_n$ |

- $E_i$ refers to **expected count** of category $i$

| Type $1$| Type $2$ |  $\ldots$ | Type $n$ |
|:-------:|:--------:|:---------:|:--------:|
|$E_1$    |   $E_2$  | $\ldots$  |    $E_n$ |
 


## Chi-squared goodness of fit test {.smaller}

**Goal:** Compare expected counts $E_i$ with observed counts $O_i$

**Method:**

- Null hypothesis: The theoretical model for expected counts is correct
- Look for evidence against the null hypothesis
- Null hypothesis is wrong
    * If distance between observed counts and expected counts is large
    * For example if
    $$
    (O_i - E_i)^2 \geq c
    $$
    for some chosen constant $c$



## Chi-squared statistic {.smaller}

::: Definition

Global distance between observed and expected counts is defined as
$$
\chi^2 := \sum_{i=1}^n \frac{(O_i-E_i)^2}{E_i}
$$
The above is called **chi-squared statistic**

:::

**Note:** We have that 
$$
\chi^2 = 0 \qquad \iff \qquad O_i = E_i  \,\,\,\, \text{ for all } \,\,\,\, i = 1 , \, \ldots , \, n
$$


## Chi-squared statistic {.smaller}

**Remarks:**

- We expect small differences between $O_i$ and $E_i$ 
    * Therefore $\chi^2 > 0$

- However $O_i$ should not be too far away from $E_i$
    * Therefore $\chi^2$ should not be too large 

- The above imply that tests on $\chi^2$ should be one-sided





## Modelling counts {.smaller}

**Multinomial distribution:** is a model for the following experiment
  
- The experiment consists of $m$ independent trials 
- Each trial results in one of $n$ distinct possible outcomes
- The probability of the $i$-th outcome is $p_i$ on every trial, with
$$
0 \leq p_i \leq 1 \qquad \qquad \sum_{i=1}^n p_i = 1
$$
- $X_i$ counts the number of times $i$-th outcome occurred in the m trials. It holds
$$
\sum_{i=1}^n X_i = m
$$



## Modelling counts {.smaller}

**Multinomial distribution:** Schematic visualization

<br>


|  Outcome type | $1$  | $\ldots$ | $n$   | Total                   |
|:-------------:|:----:|:--------:|:-----:|:-----------------------:|
| Counts        |$X_1$ | $\ldots$ | $X_n$ | $X_1 + \ldots + X_n = m$|
| Probabilities |$p_1$ | $\ldots$ | $p_n$ | $p_1 + \ldots + p_n = 1$|





## Modelling counts {.smaller}

For $n = 2$ the multinomial reduces to a binomial:

- Each trial has $n = 2$ possible outcomes
- $X_1$ counts the number of *successes* 
- $X_2 = m − X_1$ counts the number of *failures* in $m$ trials
- Probability of *success* is $p_1$
- Probability of *failure* is $p_2 = 1 - p_1$


| Outcome types | $1$     | $2$               |
|:-------------:|:-------:|:-----------------:|
| Counts        |$X_1$    |  $X_2 = m - X_1$  |
| Probabilities |$p_1$    |  $p_2 = 1 - p_1$  |




## Multinomial distribution {.smaller}


::: Definition

Let $m,n \in \N$ and $p_1, \ldots, p_n$ numbers such that
$$
0 \leq p_i \leq 1 \,, \qquad \quad 
\sum_{i=1}^n p_i = 1
$$
The random vector $\XX = (X_1, \ldots, X_n)$ has **multinomial distribution** with $m$ trials and cell probabilities $p_1,\ldots,p_n$ if joint pmf if
$$
f (x_1, \ldots , x_n) = \frac{m!}{x_1 ! \cdot \ldots \cdot x_n !} \ p_1^{x_1} \cdot \ldots \cdot p_n^{x_n}  \,, \qquad \forall \, x_i \in \N  \, \st \sum_{i=1}^n x_i = m
$$
We denote $\XX \sim \multinomial(m,p_1,\ldots,p_n)$

:::



## Multinomial distribution {.smaller}
### Properties

Suppose that $\XX \sim \multinomial(m,p_1,\ldots,p_n)$

- Then $X_i$ is binomial with $m$ trials and probability $p_i$
- We write $X_i \sim \binomial(m,p_i)$ and the pmf is
$$
f(x_i) = P(X = x_i) = \frac{m!}{x_i! \cdot (1-x_i)!} \, p_i^{x_i} (1-p_i)^{1-x_i} 
\qquad \forall \, x_i = 0 , \ldots , m 
$$
- We have
$$
\Expect[X_i] = m p_i \qquad \qquad 
\Var[X_i] = m p_i (1-p_i)
$$



## Counts with Multinomial distribution {.smaller}

- $O_i$ refers to **observed count** of category $i$


- $E_i$ refers to **expected count** of category $i$

- We suppose that Type $i$ is observed with probability $p_i$ and
$$
0 \leq p_i \leq 1 \,, \qquad \quad p_1 + \ldots + p_n = 1
$$

- Total number of observations is $m$

- The counts are modelled by
$$
(O_1, \ldots, O_n) \sim \multinomial (m, p_1, \ldots, p_n)
$$




## Counts with Multinomial distribution {.smaller}

Therefore the observed and expected counts are
$$
O_i \sim \binomial (m,p_i) \qquad \qquad 
E_i = \Expect[ O_i ] = m p_i
$$


::: Definition

The chi-squared statistic for multinomial counts is
$$
\chi^2 = \sum_{i=1}^n \frac{(O_i-E_i)^2}{E_i} 
       = \sum_{i=1}^n \frac{( O_i - m p_i )^2}{ m p_i }
$$

:::

**Question**: What is the distribution of $\chi^2 \,$?




## Distribution of chi-squared statistic {.smaller}

::: Theorem

Suppose the counts $(O_1, \ldots, O_n) \sim \multinomial (m,p_1, \ldots, p_n)$. Then
$$
\chi^2 = \sum_{i=1}^n \frac{( O_i - m p_i )^2}{ m p_i } \ \stackrel{{\rm d}}{\longrightarrow} \ \chi_{n-1}^2
$$
when $m \to \infty$, where the convergence is in distribution
:::

- Hence the distribution of $\chi^2$ is approximately $\chi_{n-1}^2$ when $m$ is large

- The above Theorem is due to Karl Pearson in 1900 paper [link](https://www.tandfonline.com/doi/abs/10.1080/14786440009463897)

- Proof is difficult. Seven different proofs are presented in this paper [link](https://arxiv.org/abs/1808.09171)



## Distribution of chi-squared statistic {.smaller}
### Heuristic proof of Theorem


- Since $O_i \sim \binomial(m, p_i)$, the Central 
Limit Theorem implies
$$
 \frac{O_i - m p_i }{ \sqrt{m p_i(1 - p_i) } } \
 \stackrel{{\rm d}}{\longrightarrow} \ N(0,1)
$$
as $m \to \infty$

- In particular, since $(1-p_i)$ in constant, we have
$$
 \frac{O_i - m p_i }{ \sqrt{m p_i } } \ \approx \ \frac{O_i - m p_i }{ \sqrt{m p_i(1 - p_i) } } \ \approx \ N(0,1) 
$$




## Distribution of chi-squared statistic {.smaller}
### Heuristic proof of Theorem

- Squaring the above we get
$$
 \frac{(O_i - m p_i)^2 }{ m p_i } \ \approx   \ N(0,1)^2 = \chi_1^2  
$$


- If the above random variables were pairwise independent, we would obtain
$$
\chi^2 = \sum_{i=1}^n \frac{(O_i - m p_i)^2 }{ m p_i } \ \approx  \ \sum_{i=1}^n \chi_1^2 = \chi_n^2  
$$



## Distribution of chi-squared statistic {.smaller}
### Heuristic proof of Theorem

- However the $O_i$ are not independent

- This is because $O_i$ is marginal of multinomial distribution

- In particular the linear constraint 
$$
O_1 + \ldots + O_n = m
$$
holds, which leads to dependence

- A rather technical proof is needed to prove the actual thesis that
$$
\chi^2 = \sum_{i=1}^n \frac{(O_i - m p_i)^2 }{ m p_i } \ \approx \ \chi_{n-1}^2  
$$



## Quality of chi-squared approximation {.smaller}

- Define expected counts $E_i := m p_i$ 
- Consider approximation from Theorem:
$$
\chi^2 = \sum_{i=1}^n \frac{(O_i - E_i)^2 }{ E_i } \ \approx \ \chi_{n-1}^2  
$$

- The approximation is:
  
  | **Good**     |  $E_i \geq 5  \, \text{ for all } \, i = 1 , \ldots n$ |
  |:------------ |:------------------------------------------------------ |
  |**Acceptable**|              $E_i \geq 1.5 \,$ for all $\, i = 1 , \ldots n$ |
  | **Bad**      |                $E_i < 1.5 \,$ for some $\, i = 1 , \ldots n$ |
    

- **Good** or **Acceptable**: Use $\chi_{n-1}^2$ approximation of $\chi^2$
- **Bad**: Use Monte Carlo simulations (more on this later)




## The chi-squared goodness-of-fit test {.smaller}

**Setting:** 

- Population consists of items of $n$ different types
- $p_i$ is probability that an item selected at random is of type $i$
- $p_i$ is unknown and needs to be estimated

- As guess for $p_1,\ldots,p_n$ we take $p_1^0, \ldots, p_n^0$ such that
$$
0 \leq p_i^0 \leq 1 \qquad \qquad \sum_{i=1}^n p_i^0 = 1
$$



## The chi-squared goodness-of-fit test {.smaller}

**Hypothesis Test:** We test for equality of $p_i$ to $p_i^0$
\begin{align*}
& H_0 \colon p_i = p_i^0 \qquad \text{ for all } \, i = 1, \ldots, n \\
& H_1 \colon p_i \neq p_i^0 \qquad \text{ for at least one } \, i 
\end{align*}


**Sample**: 

- We draw $m$ items from population
- $O_i$ denotes the number of items of type $i$ drawn
- Therefore $(O_1, \ldots, O_n) \sim \multinomial (m,p_1, \ldots, p_n)$


## The chi-squared goodness-of-fit test {.smaller}

**Data**: Vector of counts $(o_1,\ldots,o_n)$

**Schematically:** We can represent probabilities and counts in a table

<br>

|  Type      | $1$  | $\ldots$ | $n$   | Total                   |
|:-----------:|:----:|:--------:|:-----:|:-----------------------:|
| Counts      |$o_1$ | $\ldots$ | $o_n$ | $m$|
|Probabilities|$p_1$ | $\ldots$ | $p_n$ | $1$|



## The chi-squared goodness-of-fit test {.smaller}
### Procedure

1. **Calculation**: 
    * Compute total counts and expected counts
    $$
    m = \sum_{i=1}^n o_i \qquad \quad E_i = m p_i^0
    $$ 
    * Compute the chi-squared statistic
    $$
    \chi^2 = \sum_{i=1}^n \frac{ (o_i - E_i)^2 }{E_i}
    $$



## The chi-squared goodness-of-fit test {.smaller}
### Procedure

2. **Statistical Tables or R**: 
    * Check that $E_i \geq 1.5$ for all $i = 1, \ldots, n$
    * In this case $\chi^2 \ \approx \ \chi_{n-1}^2$
    * Find critical value in chi-squared [Table 13.5](files/Statistics_Tables.pdf)
    $$
    \chi^2_{n - 1} (0.05)
    $$
    * Alternatively, compute p-value in R
    $$
    p := P( \chi^2_{n - 1} > \chi^2 )
    $$




## The chi-squared goodness-of-fit test {.smaller}
### Procedure

3. **Interpretation**:
    * Reject $H_0$ if 
    $$
    \chi^2 > \chi_{n - 1}^2 (0.05)  \qquad \text{ or } \qquad p < 0.05
    $$
    * Do not reject $H_0$ if 
    $$
    \chi^2 \leq \chi^2_{n - 1} (0.05)  \qquad \text{ or } \qquad 
    p \geq 0.05
    $$





## The chi-squared goodness-of-fit test in R {.smaller}
### General commands

1. Store the counts $o_1,\ldots, o_n$ in R vector
    * ``counts <- c(o1, ..., on)``

2. Store the null hypothesis probabilities $p_1^0,\ldots, p_n^0$ in R vector
    * ``null_probs <- c(p1, ..., pn)``

3. Perform a goodness-of-fit test on ``counts`` with ``null_probs``
    * ``chisq.test(counts, p = null_probs)``

4. Read output
    * Output is similar to t-test and F-test
    * The main quantity of interest is p-value




## Comments on ``chisq.test`` {.smaller}


1. R performs a goodness-of-fit test on ``counts`` with ``null_probs`` with
    * ``chisq.test(counts, p = null_probs)``

2. R implicitly assumes the null hypothesis is 
$$
H_0 \colon p_i = p_i^0   \qquad \text{ for all } \, i = 1 , \ldots, n
$$

3. By default R computes p-value using $\chi_{n-1}^2$ approximation

4. R can compute p-value with Monte Carlo simulation using option
    * ``simulate.p.value = TRUE``



# Part 3: <br>Goodness-of-fit test:<br> Examples {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Example: Blood counts {.smaller}

- Suppose to have counts of blood type for some people

    | A  |  B  |  AB  |  O  |
    |:--:|:---:|:----:|:---:|
    |2162| 738 | 228  |2876 |

- We want to compare the above to the probability model

    | A  |  B  |  AB  |  O  |
    |:--:|:---:|:----:|:---:|
    |1/3 | 1/8 |  1/24| 1/2 |


- Hence the null hypothesis probabilities are 
$$
p_1^0 = \frac{1}{3} \qquad \quad   p_2^0 = \frac{1}{8} \qquad \quad
p_3^0 = \frac{1}{24} \qquad \quad p_4^0 = \frac{1}{2}
$$



## Example: Blood counts {.smaller}
### Goodness-of-fit test by hand

1. **Calculation**: 
    * Compute total counts
    $$
    m = \sum_{i=1}^n o_i = 2162 + 738 + 228 + 2876 = 6004
    $$
    
    
    
## Example: Blood counts {.smaller}
### Goodness-of-fit test by hand

1. **Calculation**:     
    * Compute expected counts
    \begin{align*}
    E_1 & = m p_1^0 = 6004 \times \frac{1}{3} = 2001.3  \\ 
    E_2 & = m p_2^0 = 6004 \times \frac{1}{8} = 750.5   \\
    E_3 & = m p_3^0 = 6004 \times \frac{1}{24} = 250.2   \\
    E_4 & = m p_4^0 = 6004 \times \frac{1}{2} = 3002  
    \end{align*}


## Example: Blood counts {.smaller}
### Goodness-of-fit test by hand

1. **Calculation**: 
    * Compute the chi-squared statistic
    \begin{align*}
    \chi^2 & = \sum_{i=1}^n \frac{ (o_i - E_i)^2 }{E_i} \\
           & = \frac{ (2162 − 2001.3)^2 }{ 2001.3 } + 
               \frac{ (738 − 750.5)^2 }{ 750.5 } \\
           & \phantom{ = } + \frac{ (228 − 250.2)^2 }{ 250.2 } + 
               \frac{ (2876 − 3002)^2 }{ 3002 } \\    
           & = 20.37  
    \end{align*}





## Example: Blood counts {.smaller}
### Goodness-of-fit test by hand

2. **Statistical Tables**: 
    * Degrees of freedom are $\, {\rm df} = n - 1 = 3$
    * We have computed
    $$
    E_1 = 2001.3 \qquad E_2 = 750.5 \qquad E_3 = 250.2 \qquad E_4 = 3002
    $$
    * Hence $E_i \geq 5$ for all $i = 1, \ldots, n$
    * In this case $\chi^2 \ \approx \ \chi_{3}^2$
    * In chi-squared [Table 13.5](files/Statistics_Tables.pdf) we find critical value
    $$
    \chi^2_{3} (0.05) = 7.81
    $$
    



## Example: Blood counts {.smaller}
### Goodness-of-fit test by hand


3. **Interpretation:**
    * We have that
    $$
    \chi^2 = 20.37  > 7.81 = \chi_{3}^2 (0.05)
    $$
    * Therefore we reject $H_0$

4. **Conclusion:**
    * Observed counts suggest at least one of null probabilities $p_i^0$ is wrong





## Example: Fair die {.smaller}
### Goodness-of-fit test in R

- A die with $6$ faces is rolled $100$ times
- Counts observed are

| $1$| $2$ | $3$  | $4$ |  $5$|  $6$|
|:--:|:---:|:----:|:---:|:---:|:---:|
|13  | 17  | 9    | 17  | 18  | 26  |


- **Exercise**: Is the die fair?
    * Formulate appropriate goodness-of-fit test
    * Implement this test in R




## Example: Fair die {.smaller}
### Solution

- The die is fair if probability of rolling $i$ is $1/6$
- Therefore we test the following hypothesis
\begin{align*}
& H_0 \colon p_i = \frac{1}{6} \qquad \text{ for all } \, i = 1, \ldots, n \\
& H_1 \colon p_i \neq \frac{1}{6} \qquad \text{ for at least one } \, i 
\end{align*}


## Example: Fair die {.smaller}
### Solution

- Test can be performed using the code below
- Code can be downloaded here [good_fit.R](codes/good_fit.R)

```r
# Enter counts and null hypothesis probabilities
counts <- c(13, 17, 9, 17, 18, 26)
null_probs <- rep(1/6, 6)

# Perform goodness-of-fit test
ans <- chisq.test(counts, p = null_probs)

# Print answer
print(ans)
```



## Example: Fair die {.smaller}
### Solution

- Running the code in previous slide we obtain


```{r}
# Enter counts and null hypothesis probabilities
counts <- c(13, 17, 9, 17, 18, 26)
null_probs <- rep(1/6, 6)

# Perform goodness-of-fit test
ans <- chisq.test(counts, p = null_probs)

# Print answer
print(ans)
```

<br>

::: {.column width="49%"}

- Chi-squared statistic is $\, \chi^2 = 9.68$

- Degrees of freedom are $\, {\rm df} = 5$

- p-value is $p \approx 0.08$

:::


::: {.column width="35%"}

- Therefore $p > 0.05$ 

- We cannot reject $H_0$

- The die appears to be fair

:::





## Example: Voting data {.smaller}
### Goodness-of-fit test in R

- Assume there are two candidates: Republican and Democrat
- Voter can choose one of these or be undecided
- 100 people are surveyed and the results are 

| Republican | Democrat | Undecided |
|:----------:|:--------:|:---------:|
|   35       |   40     |   25      |




## Example: Voting data {.smaller}
### Goodness-of-fit test in R

- Hystorical data suggest that undecided voters are $30\%$ of population
- **Exercise**: Is difference between Republican and Democratic significant?
    * Formulate appropriate goodness-of-fit test
    * Implement this test in R
    * You are not allowed to use ``chisq.test``



## Example: Voting data {.smaller}
### Solution

- Hystorical data suggest that undecided voters are $30\%$ of population
- Therefore we can assume that probability of undecided voter is
$$
p_3^0 = 0.3
$$
- Want to test if there is difference between Republican and Democrat
- Hence null hypothesis is
$$
p_1^0 = p_2^0
$$



## Example: Voting data {.smaller}
### Solution

- Recall that
$$
p_1^0 + p_2^0 + p_3^0 = 1
$$

- Hence we deduce 
$$
p_1^0 = 0.35 \qquad \quad  p_2^0 = 0.35 \qquad \quad p_3^0 = 0.3
$$


- We test the following hypothesis
\begin{align*}
& H_0 \colon p_i = p_i^0 \qquad \text{ for all } \, i = 1, \ldots, n \\
& H_1 \colon p_i \neq p_i^0 \qquad \text{ for at least one } \, i 
\end{align*}





## Example: Voting data {.smaller}
### Solution

- We want to perform goodness-of-fit test without using ``chisq.test``

- First we enter the data

```r
# Enter counts and null hypothesis probabilities
counts <- c(35, 40 , 25)
null_probs <- c(0.35, 0.35, 0.3)
```


## Example: Voting data {.smaller}
### Solution

- Compute the total number of counts $m = o_1 + \ldots + o_n$

```r
# Compute total counts
m <- sum(counts)
```

<br>

- Compute degrees of freedom $\, {\rm df} = n - 1$

```r
# Compute degrees of freedom
degrees <- length(counts) - 1
```




## Example: Voting data {.smaller}
### Solution


- Compute the expected counts $E_i = m p_i^0$

```r
# Compute expected counts
exp_counts <- m * null_probs
```



- Compute chi-squared statistic 
$$
\chi^2 = \sum_{i=1}^n \frac{( o_i - E_i )^2}{E_i}
$$



```r
# Compute chi-squared statistic
chi_squared <- sum( (counts - exp_counts)^2 / exp_counts )
```



## Example: Voting data {.smaller}
### Solution

- Finally compute the p-value
$$
p = P( \chi_{n-1}^2 > \chi^2 ) = 1 - P( \chi_{n-1}^2 \leq \chi^2 )
$$

```r
# Compute p-value
p_value <- 1 - pchisq(chi_squared, df = degrees)

# Print p-value
cat("The p-value is:", p_value)
```

<br>

- The full code can be downloaded here [good_fit_first_principles.R](codes/good_fit_first_principles.R)


## Example: Voting data {.smaller}
### Solution


- Running the code gives the following output

```{r}
# Enter counts and null hypothesis probabilities
counts <- c(35, 40 , 25)
null_probs <- c(0.35, 0.35, 0.3)

# Compute total counts
m <- sum(counts)

# Compute degrees of freedom
degrees <- length(counts) - 1

# Compute expected counts
exp_counts <- m * null_probs

# Compute chi-squared statistic
chi_squared <- sum( (counts - exp_counts)^2 / exp_counts )

# Compute p-value
p_value <- 1 - pchisq(chi_squared, df = degrees)

# Print p-value
cat("The p-value is:", p_value)
```

<br>

- Therefore p-value is $p > 0.05$

- We do not reject $H_0$

- There is no reason to believe that Republican and Democrat are not tie




# Part 4: <br>Monte Carlo <br> simulations {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## Monte Carlo simulations {.smaller}

- Recall the approximation
$$
\chi^2 = \sum_{i=1}^n \frac{(O_i - E_i)^2 }{ E_i } \ \approx \ \chi_{n-1}^2  
$$

- Define expected counts $E_i := m p_i$ 

- We said that
    * $E_i < 5$ for some $i \quad \implies \quad$ approximation of $\chi^2$ with $\chi_{n-1}^2$ is bad
    * In this case we use **Monte Carlo simulations**

<br>

- **Question:** What is a Monte Carlo simulation?





## Monte Carlo methods {.smaller}


- **Monte Carlo methods:** 
    * Broad class of computational algorithms 
    * They rely on repeated random sampling to obtain numerical results
    * Principle: use randomness to solve problems which are deterministic

- **Why the name?**
    * Method was developed by [Stanislaw Ulam](https://en.wikipedia.org/wiki/Stanisław_Ulam)
    * His uncle liked to gamble in the Monte--Carlo Casino in Monaco





## Monte Carlo simulation: Example{.smaller}
### Approximating $\pi$ 





## Monte Carlo simulation: Example{.smaller}
### Approximating $\pi$ 


- We can approximate $\pi$ via Monte Carlo simulation

- Idea: 
    * Throw random points inside unit square
    * Count proportion of points falling inside unit circle
    * Such proportion approximates the area of the circle
    * Area of unit circle is $\pi$


## Monte Carlo simulation: Example{.smaller}
### Approximating $\pi$ 

In practice, we do the following:

- Draw $x_1, \ldots, x_N, y_1, \ldots, y_N$ from ${\rm Uniform(-1,1)}$
- Draw $y_1, \ldots, y_N$ from ${\rm Uniform(-1,1)}$
- Count the number of points $(x_i,y_i)$ falling inside circle of radius $1$
- These are points satisfying condition
$$
x_i^2 + y_i^2 \leq 1
$$
- Can estimate area of circle with proportion of points falling inside circle:
$$
\text{Area} \ = \ \pi \ \approx \ \frac{\text{Number of points } (x_i,y_i) \text{ inside circle}}{N}
$$





## Monte {.smaller}
### Plot of 1000 random points uniformly distributed in unit square

```{r}
#| fig-asp: 1

R <- 1
niters <- 1000
total <- 0
in_circle <- 0

# Function to plot the points
plot_points <- function(x, y, R) {
  plot(x, y, col = ifelse(x^2 + y^2 <= R^2, "blue", "red"), pch = 16, xlab = "X", ylab = "Y", asp = 1, frame.plot = TRUE, xlim = c(-1,1), ylim = c(-1,1) )
  points(0, 0, col = "green", pch = 3)  # Plotting the center of the circle
  legend("bottomright", legend = c("Inside Circle", "Outside Circle"), col = c("blue", "red"), pch = 16, cex = 2)
  circle <- seq(0, 2 * pi, length.out = 100)
  lines(R * cos(circle), R * sin(circle))  # Plotting the circle
}

x <- runif(niters, -R, R)
y <- runif(niters, -R, R)
plot_points(x, y, R)

```



## Monte carlo {.smaller}
### Implementation in R

```r
niters <- 10000
total <- 0
in_circle <- 0

for (j in 1:10) {
  for (i in 1:niters) {
    x <- runif(1, -1, 1); y <- runif(1, -1, 1);
    if (x^2 + y^2 <= 1) {
      in_circle <- in_circle + 1
    }
    total <- total + 1
  }
  
  area <- in_circle / total
  cat(sprintf("After %8d iterations area is %.08f, error is %.08f\n",
     (j * niters), area, abs(area - pi)))
}

```




## Monte carlo {.smaller}
### Implementation in R: Output

```{r}

R <- 1
niters <- 10000
total <- 0
in_circle <- 0

for (j in 1:10) {
  for (i in 1:niters) {
    x <- runif(1, -R, R)
    y <- runif(1, -R, R)
    if (x^2 + y^2 <= R^2) {
      in_circle <- in_circle + 1
    }
    total <- total + 1
  }
  
  area <- (2 * R)^2 * in_circle / total
  expected <- pi * R^2
   cat(sprintf("After %8d iterations area is %.08f, error is %.08f\n", (j * niters), area, abs(area - expected)))
}

```

