---
title: "Statistical Models"
subtitle: "Lecture 5"
from: markdown+emoji
author: 
    - name: Dr. Silvio Fanzon
      id: sf
      email: S.Fanzon@hull.ac.uk
      url: https://www.silviofanzon.com
      affiliations: University of Hull
    - name: Dr. John Fry
      id: jf
      email: J.M.Fry@hull.ac.uk
      url: https://www.hull.ac.uk/staff-directory/john-fry
      affiliations: University of Hull
---



::: {.content-hidden}
$
{{< include macros.tex >}}
$
:::




# Lecture 5: <br>Hypothesis tests in R <br>Part 2 {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Outline of Lecture 5

1. Functions in R
2. More Vectors
3. Data Management
4. R Style Guide
5. F-test







# Part 1: <br>Functions in R {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Expressions and objects {.smaller}

- Basic way to interact with R is through **expression evaluation**:
  * You enter an epression
  * The system evaluates it and prints the result

- Expressions work on **objects**
- **Object**: anything that can be assigned to a variable
- Objects encountered so far are:
  * Scalars
  * Vectors



## Functions and arguments {.smaller}

- **Functions** are a class of objects

- Format of a function is **name** followed by parentheses containing **arguments**

- Functions take **arguments** and **return** a result

- We already encountered several built in functions:
  * ``plot(x, y)``
  * ``lines(x, y)``
  * ``seq(x)``
  * ``print("Stats is great!")``
  * ``cat("R is great!")``
  * ``mean(x)``
  * ``sin(x)``


## Functions and arguments {.smaller}

- Functions have **actual arguments** and **formal arguments**
- Example: 
  * ``plot(x, y)`` has formal arguments two vectors ``x`` and ``y``
  * ``plot(height, weight)`` has actual arguments ``height`` and ``weight``

- When you write ``plot(height, weight)`` the arguments are matched:
  * ``height`` corresponds to x-variable
  * ``weight`` corresponds to y-variable
  * This is called **positional matching**
  
  

## Functions and arguments {.smaller}

- If a function has a lot of arguments, positional matching is tedious

- For example ``plot()`` accepts the following (and more!) arguments

| Argument |                        Description    |
|:--------:|:-------------------------------------:|
| ``x``    | x coordinate of points in the plot    |
| ``y``    | y coordinate of points in the plot    |
| ``type`` | Type of plot to be drawn              |
| ``main`` | Title of the plot                     |
| ``xlab`` | Label of x axis                       |
| ``ylab`` | Label of y axis                       | 
| ``pch``  | Shape of points                       |





## Functions and arguments {.smaller}

Issue with having too many arguments is the following:

- We might want to specify ``pch = 2``
- But then we would have to match all the arguments preceding
``pch``
  * ``x``
  * ``y``
  * ``type``
  * ``xlab``
  * ``ylab``



## Functions and arguments {.smaller}

- Thankfully we can use **named actual arguments**:
  * The name of a formal argument can be matched to an actual argument
  * This is independent of position

- For example we can specify ``pch = 2`` by the call
  * ``plot(weight, height, pch = 2)``

- In the above:
  * ``weight`` is implicitly matched to ``x``
  * ``height`` is implicitly matched to ``y``
  * ``pch`` is explicitly matched to ``2``
  
- Note that the following call would give same output
  * ``plot(x = weight, y = height, pch = 2)``






## Functions and arguments {.smaller}

- Named actual arguments **override** positional arguments
- Example: The following commands yield the same plot
  * ``plot(height, weight)``
  * ``plot(x = height, y = weight)``
  * ``plot(y = weight, x = height)``





## Functions and arguments {.smaller}

We have already seen another example of **named actual arguments**

- ``seq(from = 1, to = 11, by = 2)``
- ``seq(1, 11, 2)``
- These yield the same output. Why?
- Because in this case **named actual arguments** match **positional arguments**




## Functions and arguments {.smaller}

If however we want to divide the interval $[1, 11]$ in $5$ equal parts:

- Have to use ``seq(1, 11, length.out = 6)``
- This is different from ``seq(1, 11, 6)``

```{r}
#| echo: true
x <- seq(1, 11, length.out = 6)
print(x)
y <- seq(1, 11, 6)
print(y)
```


- This is because:
  * The 3rd positional argument of ``seq()`` is ``by``
  * The 4th positional argument is ``length.out``



## Functions and arguments {.smaller}
### Warning

- You can call functions without specifying arguments
- However you have to use brackets ``()``
- Example:
  * ``getwd()`` -- which outputs current working directory
  * ``ls()`` -- which outputs names of objects currently in memory






# Part 2: <br>More Vectors{background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## More vectors {.smaller}

- Vectors of numbers are objects
- Further type of vectors are:
  * Character vectors
  * Logical vectors


## Character vectors {.smaller}

- A character vector is a vector of text strings
- Elements are specified and printed in quotes

```{r}
#| echo: true
x <- c("Red", "Green", "Blue")
print(x)
```

- You can use single- or double-quote symbols to specify strings
- This is as long as the left quote is the same as the right quote

```{r}
#| echo: true
x <- c('Red', 'Green', 'Blue')
print(x)
```



## Character vectors {.smaller}

``Print`` and ``cat`` produce different output on character vectors:

- ``print(x)`` prints all the strings in ``x`` separately
- ``cat(x)`` concatenates strings. There is no way to tell how many were there


::: {.column width="45%"}

```r
x <- c("Red", "Green", "Blue")
print(x)
cat(x)
```

```{r}
x <- c("Red", "Green", "Blue")
print(x)
cat(x)
```

:::



::: {.column width="45%"}

```r
y <- c("Red Green", "Blue")
print(y)
cat(y)
```

```{r}
y <- c("Red Green", "Blue")
print(y)
cat(y)
```

:::


## Logical vectors {.smaller}

- Logical vectors can take the values ``TRUE``, ``FALSE`` or ``NA``
- ``TRUE`` and ``FALSE`` can be abbreviated with ``T`` and ``F``

```{r}
#| echo: true
x <- c(T,T,F,T)
print(x)
```


## Logical vectors {.smaller}

- Logical vectors are extremely useful to evaluate conditions

- Example:
  * given a numerical vector ``x``
  * we want to count how many entries are above a value ``t``


```{r}
#| echo: true
x <- seq(from = 1 , to = 10, by = 1)
print(x)

y <- ( x > 5 )
print(y)
```


## Logical vectors {.smaller}
### Application

- Generate a vector of 1000 numbers from $N(0,1)$
- Count how many entries are above the mean $0$
- Since there are many (1000) entries, we expect a result close to 500

**Question**: How to do this?

**Hint**: When you sum ``T`` and ``F`` they are interpreted as ``1`` and ``0``

```{r}
#| echo: true
T + T
T + F
F + F
```


## Logical vectors {.smaller}
### Application


- The function ``sum(x)`` sums the entries of a vector ``x``
- We can use ``sum(x)`` to count the number of ``T`` entries in a logical vector ``x``



```r
# Generate vector with 1000 normal entries
x <- rnorm(1000)

# Generate logical vector of entries above 0
y <- (x > 0)

# Count entries above zero
above_zero <- sum(y)
cat("Number of entries above average 0 is", above_zero)
cat("This is pretty close to 500!")
```


```{r}
# Generate vector with 1000 normal entries
x <- rnorm(1000)

# Generate logical vector of entries above 0
y <- (x > 0)

# Count entries above zero
above_zero <- sum(y)

cat("Number of entries above average 0 is", above_zero)
cat("This is pretty close to 500!")
```


## Missing values {.smaller}

- In practical data analysis, a data point is frequently unavailable
- Statistical software needs ways to deal with this
- R allows vectors to contain a special ``NA`` value - Not Available
- ``NA`` is carried through in computations: operations on ``NA`` yield ``NA`` as the result


```{r}
#| echo: true

2 * NA
NA + NA
T + NA
```




## Functions that create vectors {.smaller}

The main functions to generate vectors are

- ``c()`` concatenate
- ``seq()`` sequence
- ``rep()`` replicate

We have already met ``c()`` and ``seq()`` but there are more details to discuss




## Functions that create vectors {.smaller}
### Concatenate

- ``c()`` generates a vector containing the input values

```{r}
#| echo: true
x <- c(1, 2, 3, 4, 5)
print(x)
```


- ``c()`` can also concatenate vectors 

```{r}
#| echo: true
x <- c(1, 2, 3, 4, 5)
y <- c(6, 7, 8)

z <- c(x, y, 9)
print(z)
```



## Functions that create vectors {.smaller}
### Concatenate


- You can assign names to vector elements

- This modiï¬es the way the vector is printed

```{r}
#| echo: true

x <- c(first = "Red", second = "Green", third = "Blue")
print(x)
```

- The names can be extracted with ``names(x)``
- The values can be extracted with ``unname(x)``

```r
print( names(x) )
print( unname(x) )
```


```{r}
x <- c(first = "Red", second = "Green", third = "Blue")
print( names(x) )
print( unname(x) )
```


## Functions that create vectors {.smaller}
### Concatenate

- All elements of a vector have the **same** type
- Concatenating vectors of different types leads to conversion


```{r}
#| echo: true

c(FALSE, 2)
```


```{r}
#| echo: true

c(pi, "stats")

```


```{r}
#| echo: true

c(TRUE, "stats")
```




## Functions that create vectors {.smaller}
### Sequence

- Recall the syntax of ``seq`` is
  * ``seq(from =, to =, by =, length.out =)``

- Omitting the third argument assumes that ``by = 1``

```{r}
#| echo: true

x <- seq(1, 6)
print(x)
```


- The same output of ``seq(x1, x2)`` is obtained with ``x1:x2``

```{r}
#| echo: true

y <- 1:6
print(y)
```

- Syntax ``x1:x2`` is **preferred** to ``seq(x1, x2)``




## Functions that create vectors {.smaller}
### Replicate


``rep`` generates repeated values from a vector:

- ``x`` vector 
- ``n`` integer
- Then ``rep(x, n)`` repeats vector ``x`` n times

```{r}
#| echo: true

x <- c(2, 1, 3)
y <- rep(x, 4)

print(y)
```
  

## Functions that create vectors {.smaller}
### Replicate


The second argument of ``rep()`` can also be a vector:

- ``x`` vector
- ``y`` vector
- ``rep(x, y)`` repeats entries of ``x`` as many times as corresponding entries of ``y``


```{r}
#| echo: true

x <- c(2, 1, 3)
y <- c(1, 2, 3)

z <- rep(x, y)

print(z)
```


## Functions that create vectors {.smaller}
### Replicate

- ``rep`` can be useful to create a vector of labels
- **Example**: Suppose we want to collect some numeric data on 3 Cats and 4 Dogs


```{r}
#| echo: true

x <- c("Cat", "Dog")
y <- rep(x, c(3, 4))

print(y)
```








# Part 3: <br>Data Management {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::










# Part 4: <br>R Style Guide {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::


## R Style Guide {.smaller}

- Styling your code is optional
- However it is considered good manners to do so
- Good coding style makes code more readable
- Highly recommended, especially for assignments
- The next few slides on Style are based on these two posts:
  * Style Guide by Hadley Wickham ([link](http://adv-r.had.co.nz/Style.html#:~:text=When%20indenting%20your%20code%2C%20use,definition%20runs%20over%20multiple%20lines.))
  * Google's R Style Guide ([link](https://google.github.io/styleguide/Rguide.html))



## R Style guide {.smaller}
### File names

They should be meaningful and end in .R

```r
# Good
football-models.R  
utility-functions.R
homework_1.R
homework1.R

# Bad
footballmodels.r # Hard to read
stuff.r          # What is inside this file?
code.r           # Same as above

```


## R Style guide {.smaller}
### Objects names

- Objects names shoulde be lowercase
- Use an underscore (``_``) to separate words within a name
- Variable names should be nouns, not verbs
- Come up with names that are concise and meaningful

```r
# Good
day_one  # This will clearly store the value of first day
day_1    # Still clear


# Bad
first_day_of_the_month  # Too long
dayone                  # Hard to read
DayOne                  # Mix of upper and lower case
fdm                     # Hard to guess what this means
```




## R Style guide {.smaller}
### Functions names

- Name functions with ``BigCamelCase`` ([link](https://en.wikipedia.org/wiki/Camel_case))
- This is to clearly distinguish functions from other objects
- Functions names should be verbs
- Come up with names that are concise and meaningful

```r
# Good
DoNothing <- function() {
  return(invisible(NULL))
}

# Bad
donothing <- function() {
  return(invisible(NULL))
}
```




## R Style guide {.smaller}
### Object and functions names

If possible avoid using names of existing functions and variables


```r
# Bad
T <- FALSE                  # T is reserved for the boolean TRUE
c <- 10                     # c denotes the concatenation operator
mean <- function(x) sum(x)  # mean already denotes a built in function
```



## R Style guide {.smaller}
### Assignment

Use ``<-`` and not ``=`` for assignment

```r 
# Good
x <- 5

# Bad
x = 5
```




## R Style guide {.smaller}
### Spacing

- Spacing is **really** something you should be careful about
- Place spaces around all infix operators (``=``, ``+``, ``-``, ``<-``, etc.) 
- Place spaces around ``=`` when calling a function 
- Always put a space after a comma, never before (like in regular English)

```r 
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
```




## R Style guide {.smaller}
### Spacing with Brackets

- Do not place spaces around code in parentheses or square brackets
- Unless there is a comma

```r
# Good
if (condition) do(x)
diamonds[5, ]

# Bad
if ( condition ) do(x)  # No spaces around condition
x[1,]                   # Needs a space after the comma
x[1 ,]                  # Space goes after comma not before
```




## R Style guide {.smaller}
### Spacing - Exceptions

- Symbols ``:``, ``::`` and ``:::`` do not need spacing

```r
# Good
x <- 1:10

# Bad
x <- 1 : 10
```

- Place a space before left parentheses, except in a function call

```r
#Good
if (condition) do(x)
plot(x, y)

# Bad
if(condition)do(x)    # (condition) needs spacing
plot (x, y)           # This does not need spacing
```



## R Style guide {.smaller}
### Extra Spacing

Extra spacing is ok if it improves alignment of ``=`` or ``<-``

```r
list(
  total = a + b + c, 
  mean  = (a + b + c) / n
)
```



## R Style guide {.smaller}
### Curly braces

- An opening curly brace should never go on its own line 
- An opening curly brace should always be followed by a new line
- Always **indent** the code inside curly braces


::: {.column width="45%"}

```r
# Good

if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} 
```

:::


::: {.column width="45%"}

```r
# Bad

if (y < 0 && debug)
message("Y is negative")


if (y == 0) 
{
  log(x)} 
```

:::





## R Style guide {.smaller}
### Line length

- Limit code to 80 characters per line
- This fits comfortably on a printed page 
- If you run out of room, encapsulate some of the work in separate function




## R Style guide {.smaller}
### Indentation

- When indenting your code, use two spaces
- Never use tabs or mix tabs and spaces
- Indentation should be used for **functions**, **if**, **for**, etc.

```r
SumTwoNumbers <- function(x, y) {
  s = x + y
  return(s)
}
```



## R Style guide {.smaller}
### Indentation - Exception

If a function definition runs over multiple lines, indent the second line to where the definition starts

```r
long_function_name <- function(a = "a long argument", 
                               b = "another argument",
                               c = "another long argument") {
  # As usual code is indented by two spaces.
}
```



## R Style guide {.smaller}
### Named arguments

- Often you can call a function without explicitly naming arguments:
  * ``plot(height, weight)``
  * ``mean(weight)``

- This might be fine for ``plot()`` or ``mean``

- However for less common functions:
  * One might struggle to remember the meaning of **arguments positions**
  * It is therefore good practice to **name arguments**

```r
# Good
seq(from = 1, to = 11, by = 1)

# Bad
seq(1, 11, 1)
```



## R Style guide {.smaller}
### Use explicit returns

- Functions can return objects
- R has an implicit return feature
- Do not rely on this feature, but explicitly mention ``return(object)``

```r
# Good
AddValues <- function(x, y) {
  return(x + y)                     # Function returns x+y
}

# Bad
AddValues <- function(x, y) {
  x + y                             # Function still returns x+y
}                                   # but it is not immediate to see it
```



## R Style guide {.smaller}
### Comments

- **Most importantly**: Comment your code
- Each line of a comment should begin with comment symbol ``#`` and a single space

```r 
# Here we sum two numbers  
x+y
```


- Use commented lines of ``-`` and ``=`` to break up code into easily readable chunks

```r
# Load data ---------------------------

# Plot data ---------------------------
```





# Part 5: <br>F-test {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::




## References
